---
title: "02_data_exploration.Rmd"
output: html_document
date: "2026-06-26"
---

This R Markdown file reads the cleaned dataset from `01_data_cleaning.Rmd` （`data/processed/analysis_long.rds`）and
explores structure, participation patterns, missingness, and descriptive score
distributions. It does not analyse the hypotheses.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read datasets and libraries
```{r load-packages}
library(tidyverse)
library(knitr)
```

```{r}
processed_data_folder <- file.path("data", "processed")
figure_folder <- file.path("figures")
analysis_long <- readRDS(cleaned_data_path)
```

# Participation Patterns And Attrition
```{r}
participation <- analysis_long |>
  distinct(participant_id, wave_number) |>
  mutate(observed = TRUE) |>
  pivot_wider(
    names_from = wave_number,
    values_from = observed,
    names_prefix = "wave",
    values_fill = FALSE
  ) |>
  mutate(
    participation_pattern = paste0(
      as.integer(wave1),
      as.integer(wave2),
      as.integer(wave3)
    )
  )

attrition_table <- participation |>
  count(participation_pattern, name = "participants") |>
  arrange(participation_pattern)

kable(attrition_table)

#The pattern 111 means that the key appears in all three files. The pattern 100 means Wave 1 only, while 011 means Waves 2 and 3 only.
```
# Outcome Distributions
```{r}
score_data <- analysis_long |>
  select(
    wave,
    depression_score,
    anxiety_score,
    loneliness_score,
    social_contact_score,
    social_support_score,
    life_satisfaction
  ) |>
  pivot_longer(
    cols = -wave,
    names_to = "variable",
    values_to = "value"
  )

outcome_summary <- score_data |>
  group_by(wave, variable) |>
  summarise(
    n = sum(!is.na(value)),
    missing_n = sum(is.na(value)),
    mean = mean(value, na.rm = TRUE),
    standard_deviation = sd(value, na.rm = TRUE),
    minimum = min(value, na.rm = TRUE),
    p25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    p75 = quantile(value, 0.75, na.rm = TRUE),
    maximum = max(value, na.rm = TRUE),
    .groups = "drop"
  )

kable(outcome_summary, digits = 2)

```

# Time-Varying and Time-Invariant Variables
```{r}
time_invariant_check <- analysis_long |>
  group_by(participant_id) |>
  summarise(
    n_sex_values = n_distinct(sex, na.rm = TRUE),
    n_cohort_values = n_distinct(cohort, na.rm = TRUE),
    .groups = "drop"
  ) |>
  summarise(
    participants = n(),
    sex_changes_or_conflicts = sum(n_sex_values > 1),
    cohort_changes_or_conflicts = sum(n_cohort_values > 1)
  )

kable(time_invariant_check)
```

#Duplicate observations
Duplicate checks use the temporary cross-cohort participant key. A duplicate row here means that the same key appears more than once within one wave file.
```{r}
duplicate_table <- analysis_long |>
  count(wave, participant_id) |>
  group_by(wave) |>
  summarise(
    duplicate_rows = sum(n - 1),
    duplicate_ids = sum(n > 1),
    .groups = "drop"
  )

kable(duplicate_table)
```

#Missingness
Missingness is examined after negative special codes have been changed to NA. The table below focuses on variables relevant to the planned analysis.
####support have big missingness
-single imputation? if thats only around 3% (e-resources Mice)
- some of them need multiple imputation?
predict with other variable disturbation of other variables 

```{r}
missing_data <- analysis_long |>
  select(
    wave,
    depression_item1,
    depression_item2,
    anxiety_item1,
    anxiety_item2,
    loneliness_item1,
    loneliness_item2,
    loneliness_item3,
    loneliness_item4,
    support_item1,
    support_item2,
    support_item3,
    support_when_sick,
    support_for_listening,
    life_satisfaction,
    depression_score,
    anxiety_score,
    loneliness_score,
    social_contact_score,
    social_support_score
  ) |>
  pivot_longer(
    cols = -wave,
    names_to = "variable",
    values_to = "value"
  )

missingness_table <- missing_data |>
  group_by(wave, variable) |>
  summarise(
    total_n = n(),
    missing_n = sum(is.na(value)),
    missing_percent = 100 * missing_n / total_n,
    .groups = "drop"
  ) |>
  arrange(wave, desc(missing_percent))

kable(missingness_table, digits = 1)

missingness_figure <- missingness_table |>
  ggplot(
    aes(
      x = wave,
      y = missing_percent,
      fill = wave
    )
  ) +
  geom_col() +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Missingness in planned analysis variables",
    x = NULL,
    y = "Missing records (%)"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
missingness_figure
ggsave(
  filename = file.path(figure_folder, "missingness_figure.png"),
  plot = missingness_figure,
  width = 9,
  height = 7,
  dpi = 300
)

# Social-support missingness is much higher than for the mental-health scores.
# Breaking it down by cohort helps identify whether questionnaire routing or cohort eligibility is responsible.

support_missingness_by_cohort <- analysis_long |>
  group_by(wave, cohort) |>
  summarise(
    total_records = n(),
    observed_support = sum(!is.na(social_support_score)),
    missing_support = sum(is.na(social_support_score)),
    missing_percent = 100 * missing_support / total_records,
    .groups = "drop"
  )

kable(support_missingness_by_cohort, digits = 1)
##structural questionnaire routing, not ordinary missing data (not gonna impute NCDS or BCS70)
#--> exploratory for Use support_when_sick, which has coverage across all cohorts.

# The two additional support indicators have different routing patterns. This table shows their coverage without combining them into the three-item score.

support_indicator_availability <- analysis_long |>
  group_by(wave, cohort) |>
  summarise(
    total_records = n(),
    sick_observed = sum(!is.na(support_when_sick)),
    sick_missing_percent = 100 * mean(is.na(support_when_sick)),
    listen_observed = sum(!is.na(support_for_listening)),
    listen_missing_percent = 100 * mean(is.na(support_for_listening)),
    .groups = "drop"
  )

kable(support_indicator_availability, digits = 1)

```

# Listwise Deletion Check
This table acknowledges how much data would be lost if the planned model used
complete cases only. It does not create a listwise-deleted dataset. A later
mixed-effects model should use maximum likelihood estimation so available
participant-wave records can contribute under the model assumptions.
```{r}
planned_model_variables <- c(
  "depression_score",
  "loneliness_score",
  "social_support_score",
  "sex",
  "cohort"
)

complete_case_rows <- analysis_long |>
  select(all_of(planned_model_variables)) |>
  drop_na() |>
  nrow()

listwise_deletion_table <- data.frame(
  item = c(
    "Rows before complete-case filtering",
    "Rows that would remain after listwise deletion",
    "Current project decision",
    "Future modelling decision"
  ),
  value = c(
    as.character(nrow(analysis_long)),
    as.character(complete_case_rows),
    "Do not create a listwise-deleted dataset here",
    "Use maximum likelihood estimation in the mixed-effects model"
  )
)

kable(listwise_deletion_table)
```

# Descriptive Figures

```{r}
all_score_histograms <- ggplot(score_data, aes(x = value)) +
  geom_histogram(binwidth = 1, colour = "white", fill = "#3B82A0") +
  facet_grid(variable ~ wave, scales = "free") +
  labs(
    x = "Score",
    y = "Number of responses"
  ) +
  theme_minimal()
all_score_histograms 

depression_histogram <- analysis_long |>
  filter(!is.na(depression_score)) |>
  ggplot(aes(x = depression_score)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    colour = "white",
    fill = "#2F6B84"
  ) +
  facet_wrap(~ wave, nrow = 1) +
  scale_x_continuous(breaks = 0:6) +
  labs(
    title = "Distribution of provisional PHQ-2 scores",
    x = "PHQ-2 score",
    y = "Number of response records"
  ) +
  theme_minimal(base_size = 12)
depression_histogram
ggsave(
  filename = file.path(figure_folder, "depression_histogram.png"),
  plot = depression_histogram,
  width = 8,
  height = 4,
  dpi = 300
)

loneliness_histogram <- analysis_long |>
  filter(!is.na(loneliness_score)) |>
  ggplot(aes(x = loneliness_score)) +
  geom_histogram(
    binwidth = 1,
    boundary = 3.5,
    colour = "white",
    fill = "#B05A47"
  ) +
  facet_wrap(~ wave, nrow = 1) +
  scale_x_continuous(breaks = 4:12) +
  labs(
    title = "Distribution of four-item loneliness scores",
    x = "Loneliness score",
    y = "Number of response records"
  ) +
  theme_minimal(base_size = 12)
loneliness_histogram
ggsave(
  filename = file.path(figure_folder, "loneliness_histogram.png"),
  plot = loneliness_histogram,
  width = 8,
  height = 4,
  dpi = 300
)

score_boxplots <- score_data |>
  filter(
    !is.na(value),
    variable %in% c(
      "depression_score",
      "anxiety_score",
      "loneliness_score",
      "social_support_score"
    )
  ) |>
  ggplot(aes(x = wave, y = value, fill = wave)) +
  geom_boxplot(outlier.alpha = 0.15) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Score distributions by wave",
    x = NULL,
    y = "Score"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
score_boxplots 
ggsave(
  filename = file.path(figure_folder, "score_boxplots.png"),
  plot = score_boxplots,
  width = 8,
  height = 6,
  dpi = 300
)
```
