---
title: "Data exploration and cleaning for COVID-19 Cohort Waves 1-3"
output: html_document
date: "2026-06-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This rmd audits the structure and data quality of the three UCL COVID-19 survey wave files：
Wave 1 took place in May 2020. Wave 2 took place in September and October 2020. Wave 3 took place in February and March 2021. Wave 2 and Wave 3 store the completion day and month in separate variables.


##Read datasets and library

```{r}
library(tidyverse)
library(haven)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

raw_data_folder <- "/Users/leyi/Desktop/DNP/Longitudual /project"

wave1_file <- file.path(
  raw_data_folder,
  "covid-19_wave1_survey_cls.dta"
)

wave2_file <- file.path(
  raw_data_folder,
  "covid-19_wave2_survey_cls.dta"
)

wave3_file <- file.path(
  raw_data_folder,
  "covid-19_wave3_survey_cls.dta"
)

lookup_file <- file.path(
  raw_data_folder,
  "covid-19_w1-3_variable_lookup.xlsx"
)

wave1 <- read_dta(wave1_file)
wave2 <- read_dta(wave2_file)
wave3 <- read_dta(wave3_file)

variable_lookup <- read_excel(lookup_file)

```

# Dataset Structure

```{r file-summary}
file_summary <- data.frame(
  wave = c("Wave 1", "Wave 2", "Wave 3"),
  rows = c(nrow(wave1), nrow(wave2), nrow(wave3)),
  columns = c(ncol(wave1), ncol(wave2), ncol(wave3))
)

kable(file_summary)
# The current data are stored as separate wave-specific files. Each file is wide within wave: one row per participant response and many wave-prefixed variables.
```
# Participant Identifiers
The study combines several birth cohorts. The identifier used depends on the participant's cohort:
NCDSID identifies NCDS participants.
BCSID identifies BCS70 participants.
NSID identifies Next Steps participants.
MCS cohort members are identified using MCSID and CNUM00. 
MCS parents are identified using MCSID and PNUM00.
Blank character values are treated as missing in the following counts.
```{r}
identifier_table <- data.frame(
  wave = rep(c("Wave 1", "Wave 2", "Wave 3"), each = 4),
  id_variable = rep(c("NCDSID", "BCSID", "NSID", "MCSID"), times = 3),
  non_missing = c(
    sum(trimws(wave1$NCDSID) != ""),
    sum(trimws(wave1$BCSID) != ""),
    sum(trimws(wave1$NSID) != ""),
    sum(trimws(wave1$MCSID) != ""),
    sum(trimws(wave2$NCDSID) != ""),
    sum(trimws(wave2$BCSID) != ""),
    sum(trimws(wave2$NSID) != ""),
    sum(trimws(wave2$MCSID) != ""),
    sum(trimws(wave3$NCDSID) != ""),
    sum(trimws(wave3$BCSID) != ""),
    sum(trimws(wave3$NSID) != ""),
    sum(trimws(wave3$MCSID) != "")
  ),
  unique_non_missing = c(
    n_distinct(wave1$NCDSID[trimws(wave1$NCDSID) != ""]),
    n_distinct(wave1$BCSID[trimws(wave1$BCSID) != ""]),
    n_distinct(wave1$NSID[trimws(wave1$NSID) != ""]),
    n_distinct(wave1$MCSID[trimws(wave1$MCSID) != ""]),
    n_distinct(wave2$NCDSID[trimws(wave2$NCDSID) != ""]),
    n_distinct(wave2$BCSID[trimws(wave2$BCSID) != ""]),
    n_distinct(wave2$NSID[trimws(wave2$NSID) != ""]),
    n_distinct(wave2$MCSID[trimws(wave2$MCSID) != ""]),
    n_distinct(wave3$NCDSID[trimws(wave3$NCDSID) != ""]),
    n_distinct(wave3$BCSID[trimws(wave3$BCSID) != ""]),
    n_distinct(wave3$NSID[trimws(wave3$NSID) != ""]),
    n_distinct(wave3$MCSID[trimws(wave3$MCSID) != ""])
  )
)

kable(identifier_table)
```
###A temporary participant key is created in memory for audit purposes. The cohort name is included to prevent the same number in different cohorts being treated as the same person. MCS uses both household ID and person number

```{r}
wave1 <- wave1 |>
  mutate(
    participant_id = case_when(
      CW1_COHORT == 1 ~ paste0("NCDS_", NCDSID),
      CW1_COHORT == 2 ~ paste0("BCS_", BCSID),
      CW1_COHORT == 3 ~ paste0("NS_", NSID),
      CW1_COHORT == 4 ~ paste0("MCS_CM_", MCSID, "_", CW1_CNUM00),
      CW1_COHORT == 5 ~ paste0("MCS_PARENT_", MCSID, "_", CW1_PNUM00),
      TRUE ~ NA_character_
    )
  )

wave2 <- wave2 |>
  mutate(
    participant_id = case_when(
      CW2_COHORT == 1 ~ paste0("NCDS_", NCDSID),
      CW2_COHORT == 2 ~ paste0("BCS_", BCSID),
      CW2_COHORT == 3 ~ paste0("NS_", NSID),
      CW2_COHORT == 4 ~ paste0("MCS_CM_", MCSID, "_", CW2_CNUM00),
      CW2_COHORT == 5 ~ paste0("MCS_PARENT_", MCSID, "_", CW2_PNUM00),
      TRUE ~ NA_character_
    )
  )

wave3 <- wave3 |>
  mutate(
    participant_id = case_when(
      CW3_COHORT == 1 ~ paste0("NCDS_", NCDSID),
      CW3_COHORT == 2 ~ paste0("BCS_", BCSID),
      CW3_COHORT == 3 ~ paste0("NS_", NSID),
      CW3_COHORT == 4 ~ paste0("MCS_CM_", MCSID, "_", CW3_CNUM00),
      CW3_COHORT == 5 ~ paste0("MCS_PARENT_", MCSID, "_", CW3_PNUM00),
      TRUE ~ NA_character_
    )
  )
```

#Number of waves and attritions
```{r}
wave1_ids <- wave1 |>
  select(participant_id) |>
  filter(!is.na(participant_id)) |>
  distinct() |>
  mutate(wave1 = TRUE)

wave2_ids <- wave2 |>
  select(participant_id) |>
  filter(!is.na(participant_id)) |>
  distinct() |>
  mutate(wave2 = TRUE)

wave3_ids <- wave3 |>
  select(participant_id) |>
  filter(!is.na(participant_id)) |>
  distinct() |>
  mutate(wave3 = TRUE)

participation <- full_join(wave1_ids, wave2_ids, by = "participant_id")
participation <- full_join(participation, wave3_ids, by = "participant_id")

participation <- participation |>
  mutate(
    wave1 = replace_na(wave1, FALSE),
    wave2 = replace_na(wave2, FALSE),
    wave3 = replace_na(wave3, FALSE),
    participation_pattern = paste0(
      as.integer(wave1),
      as.integer(wave2),
      as.integer(wave3)
    )
  )

attrition_table <- participation |>
  count(participation_pattern, name = "participants") |>
  arrange(participation_pattern)

kable(attrition_table)

#The pattern 111 means that the key appears in all three files. The pattern 100 means Wave 1 only, while 011 means Waves 2 and 3 only.
```

#Measurement Timing
Wave 1 contains a completion-day variable. Wave 2 and Wave 3 contain separate day and month variables.
```{r}
timing_table <- data.frame(
  wave = c("Wave 1", "Wave 2", "Wave 2", "Wave 3", "Wave 3"),
  timing_variable = c(
    "CW1_ENDDATE",
    "CW2_ENDDATED",
    "CW2_ENDDATEM",
    "CW3_ENDDATED",
    "CW3_ENDDATEM"
  ),
  non_missing = c(
    sum(!is.na(wave1$CW1_ENDDATE)),
    sum(!is.na(wave2$CW2_ENDDATED)),
    sum(!is.na(wave2$CW2_ENDDATEM)),
    sum(!is.na(wave3$CW3_ENDDATED)),
    sum(!is.na(wave3$CW3_ENDDATEM))
  ),
  unique_values = c(
    n_distinct(wave1$CW1_ENDDATE, na.rm = TRUE),
    n_distinct(wave2$CW2_ENDDATED, na.rm = TRUE),
    n_distinct(wave2$CW2_ENDDATEM, na.rm = TRUE),
    n_distinct(wave3$CW3_ENDDATED, na.rm = TRUE),
    n_distinct(wave3$CW3_ENDDATEM, na.rm = TRUE)
  ),
  minimum = c(
    min(wave1$CW1_ENDDATE, na.rm = TRUE),
    min(wave2$CW2_ENDDATED, na.rm = TRUE),
    min(wave2$CW2_ENDDATEM, na.rm = TRUE),
    min(wave3$CW3_ENDDATED, na.rm = TRUE),
    min(wave3$CW3_ENDDATEM, na.rm = TRUE)
  ),
  maximum = c(
    max(wave1$CW1_ENDDATE, na.rm = TRUE),
    max(wave2$CW2_ENDDATED, na.rm = TRUE),
    max(wave2$CW2_ENDDATEM, na.rm = TRUE),
    max(wave3$CW3_ENDDATED, na.rm = TRUE),
    max(wave3$CW3_ENDDATEM, na.rm = TRUE)
  )
)

kable(timing_table)
#The Wave 2 month values should be 9 and 10. The Wave 3 month values should be 2 and 3. Wave 1 is documented as May, so CW1_ENDDATE is interpreted as the day of May rather than as a complete date.
```

#Planned Variables and Constructs
```{r}
construct_table <- data.frame(
  construct = c(
    "Depression items",
    "Anxiety items",
    "Loneliness items",
    "Life satisfaction",
    "Social contact",
    "Social support",
    "Financial strain"
  ),
  Wave_1 = c(
    "CW1_GAD2PHQ2_3, CW1_GAD2PHQ2_4",
    "CW1_GAD2PHQ2_1, CW1_GAD2PHQ2_2",
    "CW1_LONELY_1 to CW1_LONELY_4",
    "CW1_SATN",
    "CW1_SCON1 to CW1_SCON5",
    "CW1_SOCPROV_1 to CW1_SOCPROV_3, CW1_SICK, CW1_LISTEN",
    "CW1_FINANCIALMANB, CW1_FINANCIALMAND"
  ),
  Wave_2 = c(
    "CW2_GAD2PHQ2_3, CW2_GAD2PHQ2_4",
    "CW2_GAD2PHQ2_1, CW2_GAD2PHQ2_2",
    "CW2_LONELY_1 to CW2_LONELY_4",
    "CW2_SATN",
    "CW2_SCON1 to CW2_SCON5",
    "CW2_SOCPROV_1 to CW2_SOCPROV_3, CW2_SICK, CW2_LISTEN",
    "CW2_FINANCIALMANB, CW2_FINANCIALMAND"
  ),
  Wave_3 = c(
    "CW3_GAD2PHQ2_3, CW3_GAD2PHQ2_4",
    "CW3_GAD2PHQ2_1, CW3_GAD2PHQ2_2",
    "CW3_LONELY_1 to CW3_LONELY_4",
    "CW3_SATN",
    "CW3_SCON1 to CW3_SCON5",
    "CW3_SOCPROV_1 to CW3_SOCPROV_3, CW3_SICK, CW3_LISTEN",
    "CW3_FINANCIALMANB, CW3_FINANCIALMAND"
  )
)

kable(construct_table)
```
# Cleaning the planned variables
The data dictionaries show that negative values are special codes rather than real responses:
-1 means not applicable.
-8 means no information or do not know.
-9 means does not want to answer.
The small function below changes any negative code to NA. Positive responses are kept unchanged.
```{r}
remove_special_codes <- function(x) {
  x <- as.numeric(x)
  x[x < 0] <- NA
  return(x)
}
```

The variables are now selected and renamed one wave at a time. This repetition is intentional so that the source of each variable remains visible.
```{r}
wave1_analysis <- wave1 |>
  transmute(
    participant_id = participant_id,
    wave = "Wave 1",
    depression_item1 = remove_special_codes(CW1_GAD2PHQ2_3),
    depression_item2 = remove_special_codes(CW1_GAD2PHQ2_4),
    anxiety_item1 = remove_special_codes(CW1_GAD2PHQ2_1),
    anxiety_item2 = remove_special_codes(CW1_GAD2PHQ2_2),
    loneliness_item1 = remove_special_codes(CW1_LONELY_1),
    loneliness_item2 = remove_special_codes(CW1_LONELY_2),
    loneliness_item3 = remove_special_codes(CW1_LONELY_3),
    loneliness_item4 = remove_special_codes(CW1_LONELY_4),
    contact_item1 = remove_special_codes(CW1_SCON1),
    contact_item2 = remove_special_codes(CW1_SCON2),
    contact_item3 = remove_special_codes(CW1_SCON3),
    contact_item4 = remove_special_codes(CW1_SCON4),
    contact_item5 = remove_special_codes(CW1_SCON5),
    support_item1 = remove_special_codes(CW1_SOCPROV_1),
    support_item2 = remove_special_codes(CW1_SOCPROV_2),
    support_item3 = remove_special_codes(CW1_SOCPROV_3),
    support_when_sick = remove_special_codes(CW1_SICK),
    support_for_listening = remove_special_codes(CW1_LISTEN),
    life_satisfaction = remove_special_codes(CW1_SATN),
    financial_management = remove_special_codes(CW1_FINANCIALMANB),
    financial_change = remove_special_codes(CW1_FINANCIALMAND),
    sex_code = remove_special_codes(CW1_PSEX),
    cohort_code = remove_special_codes(CW1_COHORT)
  )
```

```{r}
wave2_analysis <- wave2 |>
  transmute(
    participant_id = participant_id,
    wave = "Wave 2",
    depression_item1 = remove_special_codes(CW2_GAD2PHQ2_3),
    depression_item2 = remove_special_codes(CW2_GAD2PHQ2_4),
    anxiety_item1 = remove_special_codes(CW2_GAD2PHQ2_1),
    anxiety_item2 = remove_special_codes(CW2_GAD2PHQ2_2),
    loneliness_item1 = remove_special_codes(CW2_LONELY_1),
    loneliness_item2 = remove_special_codes(CW2_LONELY_2),
    loneliness_item3 = remove_special_codes(CW2_LONELY_3),
    loneliness_item4 = remove_special_codes(CW2_LONELY_4),
    contact_item1 = remove_special_codes(CW2_SCON1),
    contact_item2 = remove_special_codes(CW2_SCON2),
    contact_item3 = remove_special_codes(CW2_SCON3),
    contact_item4 = remove_special_codes(CW2_SCON4),
    contact_item5 = remove_special_codes(CW2_SCON5),
    support_item1 = remove_special_codes(CW2_SOCPROV_1),
    support_item2 = remove_special_codes(CW2_SOCPROV_2),
    support_item3 = remove_special_codes(CW2_SOCPROV_3),
    support_when_sick = remove_special_codes(CW2_SICK),
    support_for_listening = remove_special_codes(CW2_LISTEN),
    life_satisfaction = remove_special_codes(CW2_SATN),
    financial_management = remove_special_codes(CW2_FINANCIALMANB),
    financial_change = remove_special_codes(CW2_FINANCIALMAND),
    sex_code = remove_special_codes(CW2_PSEX),
    cohort_code = remove_special_codes(CW2_COHORT)
  )
```

```{r}
wave3_analysis <- wave3 |>
  transmute(
    participant_id = participant_id,
    wave = "Wave 3",
    depression_item1 = remove_special_codes(CW3_GAD2PHQ2_3),
    depression_item2 = remove_special_codes(CW3_GAD2PHQ2_4),
    anxiety_item1 = remove_special_codes(CW3_GAD2PHQ2_1),
    anxiety_item2 = remove_special_codes(CW3_GAD2PHQ2_2),
    loneliness_item1 = remove_special_codes(CW3_LONELY_1),
    loneliness_item2 = remove_special_codes(CW3_LONELY_2),
    loneliness_item3 = remove_special_codes(CW3_LONELY_3),
    loneliness_item4 = remove_special_codes(CW3_LONELY_4),
    contact_item1 = remove_special_codes(CW3_SCON1),
    contact_item2 = remove_special_codes(CW3_SCON2),
    contact_item3 = remove_special_codes(CW3_SCON3),
    contact_item4 = remove_special_codes(CW3_SCON4),
    contact_item5 = remove_special_codes(CW3_SCON5),
    support_item1 = remove_special_codes(CW3_SOCPROV_1),
    support_item2 = remove_special_codes(CW3_SOCPROV_2),
    support_item3 = remove_special_codes(CW3_SOCPROV_3),
    support_when_sick = remove_special_codes(CW3_SICK),
    support_for_listening = remove_special_codes(CW3_LISTEN),
    life_satisfaction = remove_special_codes(CW3_SATN),
    financial_management = remove_special_codes(CW3_FINANCIALMANB),
    financial_change = remove_special_codes(CW3_FINANCIALMAND),
    sex_code = remove_special_codes(CW3_PSEX),
    cohort_code = remove_special_codes(CW3_COHORT)
  )
```

# Provisional scale scores
The PHQ-2 and GAD-2 responses are stored as 1 to 4. Subtracting 1 converts each item to the conventional 0 to 3 range.
The contact items are stored as 1 = every day and 5 = never. Subtracting each item from 6 makes a higher score represent more frequent contact.
For the three Social Provisions items, the first two are positively worded and the third is negatively worded. They are recoded so that a higher total represents more perceived support. SICK and LISTEN are kept as separate support indicators because they use a different response scale.
```{r}
wave1_analysis <- wave1_analysis |>
  mutate(
    cohort = factor(
      cohort_code,
      levels = 1:6,
      labels = c("NCDS", "BCS70", "Next Steps", "MCS cohort member", "MCS parent", "NSHD") #The dictionaries include 6 = NSHD, although no NSHD rows occur in these files. Just for completeness to have NSHD here
    ),
    sex = factor(
      sex_code,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),
    depression_score = (depression_item1 - 1) + (depression_item2 - 1),
    anxiety_score = (anxiety_item1 - 1) + (anxiety_item2 - 1),
    loneliness_score = loneliness_item1 + loneliness_item2 +
      loneliness_item3 + loneliness_item4,
    social_contact_score = (6 - contact_item1) + (6 - contact_item2) +
      (6 - contact_item3) + (6 - contact_item4) + (6 - contact_item5),
    social_support_score = (4 - support_item1) + (4 - support_item2) +
      support_item3,
    financial_strain_before = financial_management,
    financial_strain_change = 6 - financial_change
  )

wave2_analysis <- wave2_analysis |>
  mutate(
    cohort = factor(
      cohort_code,
      levels = 1:6,
      labels = c("NCDS", "BCS70", "Next Steps", "MCS cohort member", "MCS parent", "NSHD") 
    ),
    sex = factor(
      sex_code,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),
    depression_score = (depression_item1 - 1) + (depression_item2 - 1),
    anxiety_score = (anxiety_item1 - 1) + (anxiety_item2 - 1),
    loneliness_score = loneliness_item1 + loneliness_item2 +
      loneliness_item3 + loneliness_item4,
    social_contact_score = (6 - contact_item1) + (6 - contact_item2) +
      (6 - contact_item3) + (6 - contact_item4) + (6 - contact_item5),
    social_support_score = (4 - support_item1) + (4 - support_item2) +
      support_item3,
     financial_strain_before = financial_management,
    financial_strain_change = 6 - financial_change
  )

wave3_analysis <- wave3_analysis |>
  mutate(
    cohort = factor(
      cohort_code,
      levels = 1:6,
      labels = c("NCDS", "BCS70", "Next Steps", "MCS cohort member", "MCS parent", "NSHD") 
    ),
    sex = factor(
      sex_code,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),
    depression_score = (depression_item1 - 1) + (depression_item2 - 1),
    anxiety_score = (anxiety_item1 - 1) + (anxiety_item2 - 1),
    loneliness_score = loneliness_item1 + loneliness_item2 +
      loneliness_item3 + loneliness_item4,
    social_contact_score = (6 - contact_item1) + (6 - contact_item2) +
      (6 - contact_item3) + (6 - contact_item4) + (6 - contact_item5),
    social_support_score = (4 - support_item1) + (4 - support_item2) +
      support_item3,
    financial_strain_before = financial_management,
    financial_strain_change = 6 - financial_change
  )
```

# Outcome Distributions

The three cleaned wave datasets are placed underneath each other for descriptive tables. This creates a long dataset with one row per response record per wave.
```{r}
analysis_long <- bind_rows(
  wave1_analysis,
  wave2_analysis,
  wave3_analysis
)

analysis_long <- analysis_long |>
  arrange(participant_id, wave)

score_data <- analysis_long |>
  select(
    wave,
    depression_score,
    anxiety_score,
    loneliness_score,
    social_contact_score,
    social_support_score
  ) |>
  pivot_longer(
    cols = -wave,
    names_to = "variable",
    values_to = "value"
  )

outcome_summary <- score_data |>
  group_by(wave, variable) |>
  summarise(
    n = sum(!is.na(value)),
    missing_n = sum(is.na(value)),
    mean = mean(value, na.rm = TRUE),
    standard_deviation = sd(value, na.rm = TRUE),
    minimum = min(value, na.rm = TRUE),
    p25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    p75 = quantile(value, 0.75, na.rm = TRUE),
    maximum = max(value, na.rm = TRUE),
    .groups = "drop"
  )

kable(outcome_summary, digits = 2)

```
##Visualisations
```{r}
ggplot(score_data, aes(x = value)) +
  geom_histogram(binwidth = 1, colour = "white", fill = "#3B82A0") +
  facet_grid(variable ~ wave, scales = "free") +
  labs(
    x = "Score",
    y = "Number of responses"
  ) +
  theme_minimal()
```
## Descriptive figures
```{r}
depression_histogram <- analysis_long |>
  filter(!is.na(depression_score)) |>
  ggplot(aes(x = depression_score)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    colour = "white",
    fill = "#2F6B84"
  ) +
  facet_wrap(~ wave, nrow = 1) +
  scale_x_continuous(breaks = 0:6) +
  labs(
    title = "Distribution of provisional PHQ-2 scores",
    x = "PHQ-2 score",
    y = "Number of response records"
  ) +
  theme_minimal(base_size = 12)
depression_histogram

loneliness_histogram <- analysis_long |>
  filter(!is.na(loneliness_score)) |>
  ggplot(aes(x = loneliness_score)) +
  geom_histogram(
    binwidth = 1,
    boundary = 3.5,
    colour = "white",
    fill = "#B05A47"
  ) +
  facet_wrap(~ wave, nrow = 1) +
  scale_x_continuous(breaks = 4:12) +
  labs(
    title = "Distribution of four-item loneliness scores",
    x = "Loneliness score",
    y = "Number of response records"
  ) +
  theme_minimal(base_size = 12)
loneliness_histogram

score_boxplots <- score_data |>
  filter(
    !is.na(value),
    variable %in% c(
      "depression_score",
      "anxiety_score",
      "loneliness_score",
      "social_support_score"
    )
  ) |>
  ggplot(aes(x = wave, y = value, fill = wave)) +
  geom_boxplot(outlier.alpha = 0.15) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Score distributions by wave",
    x = NULL,
    y = "Score"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
score_boxplots 
```


# Time-Varying and Time-Invariant Variables
Variables beginning with CW1_, CW2_, or CW3_ are wave-specific and may
change over time. The four unprefixed cohort identifiers are intended to link records across waves. Sex and cohort appear in each wave and should be checked for agreement before being treated as time-invariant covariates.
```{r}
wave1_prefixed <- sum(grepl("^CW1_", names(wave1)))
wave1_unprefixed <- ncol(wave1) - wave1_prefixed - 1
#The subtraction of 1 removes the temporary participant_id created during this report. It is not part of the original data.

wave2_prefixed <- sum(grepl("^CW2_", names(wave2)))
wave2_unprefixed <- ncol(wave2) - wave2_prefixed - 1

wave3_prefixed <- sum(grepl("^CW3_", names(wave3)))
wave3_unprefixed <- ncol(wave3) - wave3_prefixed - 1

variable_type_table <- data.frame(
  wave = c("Wave 1", "Wave 2", "Wave 3"),
  unprefixed_variables = c(
    wave1_unprefixed,
    wave2_unprefixed,
    wave3_unprefixed
  ),
  wave_prefixed_variables = c(
    wave1_prefixed,
    wave2_prefixed,
    wave3_prefixed
  )
)

kable(variable_type_table)
```

#Duplicate observations
Duplicate checks use the temporary cross-cohort participant key. A duplicate row here means that the same key appears more than once within one wave file.
```{r}
wave1_duplicate_rows <- sum(duplicated(wave1$participant_id))
wave2_duplicate_rows <- sum(duplicated(wave2$participant_id))
wave3_duplicate_rows <- sum(duplicated(wave3$participant_id))

wave1_duplicate_ids <- wave1 |>
  count(participant_id) |>
  filter(n > 1) |>
  nrow()

wave2_duplicate_ids <- wave2 |>
  count(participant_id) |>
  filter(n > 1) |>
  nrow()

wave3_duplicate_ids <- wave3 |>
  count(participant_id) |>
  filter(n > 1) |>
  nrow()

duplicate_table <- data.frame(
  wave = c("Wave 1", "Wave 2", "Wave 3"),
  id_variable = "temporary participant_id",
  duplicate_rows = c(
    wave1_duplicate_rows,
    wave2_duplicate_rows,
    wave3_duplicate_rows
  ),
  duplicate_ids = c(
    wave1_duplicate_ids,
    wave2_duplicate_ids,
    wave3_duplicate_ids
  )
)

kable(duplicate_table)
#zero duplicate participant-wave records
```

#Missingness
Missingness is examined after negative special codes have been changed to NA. The table below focuses on variables relevant to the planned analysis.
####support and financial have big missingness
-single imputation? if thats only around 3% (e-resources Mice)
- some of them need multiple imputation?
predict with other variable disturbation of other variables 

```{r}
missing_data <- analysis_long |>
  select(
    wave,
    depression_item1,
    depression_item2,
    anxiety_item1,
    anxiety_item2,
    loneliness_item1,
    loneliness_item2,
    loneliness_item3,
    loneliness_item4,
    support_item1,
    support_item2,
    support_item3,
    support_when_sick,
    support_for_listening,
    life_satisfaction,
    financial_management,
    financial_change,
    depression_score,
    anxiety_score,
    loneliness_score,
    social_contact_score,
    social_support_score
  ) |>
  pivot_longer(
    cols = -wave,
    names_to = "variable",
    values_to = "value"
  )

missingness_table <- missing_data |>
  group_by(wave, variable) |>
  summarise(
    total_n = n(),
    missing_n = sum(is.na(value)),
    missing_percent = 100 * missing_n / total_n,
    .groups = "drop"
  ) |>
  arrange(wave, desc(missing_percent))

kable(missingness_table, digits = 1)

missingness_figure <- missingness_table |>
  ggplot(
    aes(
      x = wave,
      y = missing_percent,
      fill = wave
    )
  ) +
  geom_col() +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Missingness in planned analysis variables",
    x = NULL,
    y = "Missing records (%)"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
missingness_figure

# Social-support missingness is much higher than for the mental-health scores.
# Breaking it down by cohort helps identify whether questionnaire routing or cohort eligibility is responsible.

support_missingness_by_cohort <- analysis_long |>
  group_by(wave, cohort) |>
  summarise(
    total_records = n(),
    observed_support = sum(!is.na(social_support_score)),
    missing_support = sum(is.na(social_support_score)),
    missing_percent = 100 * missing_support / total_records,
    .groups = "drop"
  )

kable(support_missingness_by_cohort, digits = 1)
##structural questionnaire routing, not ordinary missing data (not gonna impute NCDS or BCS70)
#--> exploratory for Use support_when_sick, which has coverage across all cohorts.

# The two additional support indicators have different routing patterns. This table shows their coverage without combining them into the three-item score.

support_indicator_availability <- analysis_long |>
  group_by(wave, cohort) |>
  summarise(
    total_records = n(),
    sick_observed = sum(!is.na(support_when_sick)),
    sick_missing_percent = 100 * mean(is.na(support_when_sick)),
    listen_observed = sum(!is.na(support_for_listening)),
    listen_missing_percent = 100 * mean(is.na(support_for_listening)),
    .groups = "drop"
  )

kable(support_indicator_availability, digits = 1)

#Also, financial_strain_before is missing for: Wave 1: 4.1%; Wave 2: 57.8%; Wave 3: 80.6% --> financial_strain_change has better coverage
financial_strain_check <- analysis_long |>
  group_by(wave) |>
  summarise(
    total_n = n(),
    observed_n = sum(!is.na(financial_strain_change)),
    missing_n = sum(is.na(financial_strain_change)),
    missing_percent = 100 * missing_n / total_n,
    mean = mean(financial_strain_change, na.rm = TRUE),
    median = median(financial_strain_change, na.rm = TRUE),
    minimum = min(financial_strain_change, na.rm = TRUE),
    maximum = max(financial_strain_change, na.rm = TRUE),
    .groups = "drop"
  )

kable(financial_strain_check, digits = 1)

analysis_long <- analysis_long |>
  mutate(
    financial_strain_category = factor(
      financial_strain_change,
      levels = 1:5,
      labels = c(
        "Much better off",
        "A little better off",
        "About the same",
        "A little worse off",
        "Much worse off"
      ),
      ordered = TRUE
    )
  )

financial_strain_table <- analysis_long |>
  count(wave, financial_strain_category, .drop = FALSE) |>
  group_by(wave) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup()

kable(financial_strain_table, digits = 1)
#Financial strain was operationalised using reported financial change compared with before the COVID-19 outbreak. Higher values indicate becoming financially worse off. The pre-COVID financial-management variable was not selected because it had substantial missingness in Waves 2 and 3.
```


# Coding and range validation
The negative values observed in the original files are documented missing-value codes. This table counts them in the planned item variables before cleaning.
```{r}
wave1_planned_names <- c(
  "CW1_GAD2PHQ2_1", "CW1_GAD2PHQ2_2",
  "CW1_GAD2PHQ2_3", "CW1_GAD2PHQ2_4",
  "CW1_LONELY_1", "CW1_LONELY_2", "CW1_LONELY_3", "CW1_LONELY_4",
  "CW1_SCON1", "CW1_SCON2", "CW1_SCON3", "CW1_SCON4", "CW1_SCON5",
  "CW1_SOCPROV_1", "CW1_SOCPROV_2", "CW1_SOCPROV_3",
  "CW1_SICK", "CW1_LISTEN", "CW1_SATN",
  "CW1_FINANCIALMANB", "CW1_FINANCIALMAND"
)

wave2_planned_names <- sub("CW1_", "CW2_", wave1_planned_names)
wave3_planned_names <- sub("CW1_", "CW3_", wave1_planned_names)

wave1_negative_codes <- wave1 |>
  select(all_of(wave1_planned_names)) |>
  mutate(across(everything(), as.numeric)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  filter(value < 0) |>
  count(variable, name = "negative_codes") |>
  mutate(wave = "Wave 1")

wave2_negative_codes <- wave2 |>
  select(all_of(wave2_planned_names)) |>
  mutate(across(everything(), as.numeric)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  filter(value < 0) |>
  count(variable, name = "negative_codes") |>
  mutate(wave = "Wave 2")

wave3_negative_codes <- wave3 |>
  select(all_of(wave3_planned_names)) |>
  mutate(across(everything(), as.numeric)) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  filter(value < 0) |>
  count(variable, name = "negative_codes") |>
  mutate(wave = "Wave 3")

negative_code_table <- bind_rows(
  wave1_negative_codes,
  wave2_negative_codes,
  wave3_negative_codes
)

kable(negative_code_table)
```


# Validate the cleaned dataset
```{r}
stopifnot(nrow(analysis_long) == nrow(wave1) + nrow(wave2) + nrow(wave3))
stopifnot(all(analysis_long$wave %in% c("Wave 1", "Wave 2", "Wave 3")))
stopifnot(!any(is.na(analysis_long$participant_id)))
stopifnot(!any(duplicated(analysis_long[c("participant_id", "wave")])))
stopifnot(all(is.na(analysis_long$depression_score) |
  analysis_long$depression_score %in% 0:6))
stopifnot(all(is.na(analysis_long$anxiety_score) |
  analysis_long$anxiety_score %in% 0:6))
stopifnot(all(is.na(analysis_long$loneliness_score) |
  analysis_long$loneliness_score %in% 4:12))
stopifnot(all(is.na(analysis_long$social_contact_score) |
  analysis_long$social_contact_score %in% 5:25))
stopifnot(all(is.na(analysis_long$social_support_score) |
  analysis_long$social_support_score %in% 3:9))
stopifnot(all(is.na(analysis_long$financial_strain_change) |
        analysis_long$financial_strain_change %in% 1:5))
stopifnot(!any(analysis_long$depression_item1 < 0, na.rm = TRUE))
stopifnot(!any(analysis_long$loneliness_item1 < 0, na.rm = TRUE))
stopifnot(
  all(is.na(analysis_long$depression_item1) |
        analysis_long$depression_item1 %in% 1:4),
  all(is.na(analysis_long$depression_item2) |
        analysis_long$depression_item2 %in% 1:4),

  all(is.na(analysis_long$anxiety_item1) |
        analysis_long$anxiety_item1 %in% 1:4),
  all(is.na(analysis_long$anxiety_item2) |
        analysis_long$anxiety_item2 %in% 1:4),

  all(is.na(analysis_long$loneliness_item1) |
        analysis_long$loneliness_item1 %in% 1:3),

  all(is.na(analysis_long$financial_change) |
        analysis_long$financial_change %in% 1:5),

  all(is.na(analysis_long$financial_strain_change) |
        analysis_long$financial_strain_change %in% 1:5)
)
```




