---
title: "03_data_analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

This R Markdown file reads the dataset from `01_data_cleaning.Rmd`
(`data/processed/analysis_long.rds`) and tests whether depressive
symptoms changed across waves, and whether loneliness, social support,
and their interaction are associated with depressive symptoms over time.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read datasets and libraries

```{r load-packages}
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(knitr)
```

```{r}
processed_data_folder <- file.path("..", "data", "processed")
cleaned_data_path <- file.path(processed_data_folder, "analysis_long.rds")
analysis_long <- readRDS(cleaned_data_path)
```

# Research Questions And Hypotheses

```{r}
hypothesis_table <- data.frame(
  hypothesis = c("H1", "H2", "H3", "H4"),
  test = c(
    "Depressive symptoms differ across waves, with Wave 3 expected to be higher than Wave 1",
    "Higher loneliness is associated with higher depressive symptoms",
    "Lower perceived social support is associated with higher depressive symptoms",
    "Social support moderates the loneliness-depression association"
  ),
  model_term = c(
    "wave",
    "loneliness_c",
    "social_support_c",
    "loneliness_c:social_support_c"
  )
)

kable(hypothesis_table)
```

# Prepare Data

The mixed-effects models use participant-wave records. Wave is treated
as a categorical variable. Loneliness and social support are
mean-centred to make the main effects and interaction easier to
interpret.

```{r}
model_data <- analysis_long |>
  mutate(
    wave = factor(wave, levels = c("Wave 1", "Wave 2", "Wave 3")),
    loneliness_c = as.numeric(scale(loneliness_score, center = TRUE, scale = FALSE)),
    social_support_c = as.numeric(scale(social_support_score, center = TRUE, scale = FALSE))
  ) |>
  select(
    participant_id,
    depression_score,
    wave,
    loneliness_c,
    social_support_c,
    sex,
    cohort
  ) |>
  drop_na()

model_sample_table <- data.frame(
  rows_in_cleaned_data = nrow(analysis_long),
  rows_in_model_data = nrow(model_data),
  participants_in_model_data = n_distinct(model_data$participant_id)
)

kable(model_sample_table)
```

# Random-Intercept Null Model

The null model contains only a random intercept for participant. This
checks whether depressive symptoms cluster within participants across
waves.

```{r}
null_model <- lmer(
  depression_score ~ 1 + (1 | participant_id),
  data = model_data,
  REML = FALSE
)

summary(null_model)
```

# Intraclass Correlation

The ICC estimates how much of the variation in depressive symptoms is
due to differences between participants.

```{r}
variance_components <- as.data.frame(VarCorr(null_model))

between_participant_variance <- variance_components |>
  filter(grp == "participant_id") |>
  pull(vcov)

residual_variance <- variance_components |>
  filter(grp == "Residual") |>
  pull(vcov)

icc_value <- between_participant_variance /
  (between_participant_variance + residual_variance)

icc_table <- data.frame(
  between_participant_variance = between_participant_variance,
  residual_variance = residual_variance,
  icc = icc_value
)

kable(icc_table, digits = 3)
```

# Build Mixed-Effects Models

Models are built incrementally. Model 1 tests wave differences. Model 2
adds loneliness and social support. Model 3 adds the loneliness by
social support interaction.

```{r}
model_1 <- lmer(
  depression_score ~ wave + sex + cohort + (1 | participant_id),
  data = model_data,
  REML = FALSE
)

model_2 <- lmer(
  depression_score ~ wave + loneliness_c + social_support_c + sex + cohort +
    (1 | participant_id),
  data = model_data,
  REML = FALSE
)

model_3 <- lmer(
  depression_score ~ wave + loneliness_c * social_support_c + sex + cohort +
    (1 | participant_id),
  data = model_data,
  REML = FALSE
)
```

# Compare Model Fit

The deviance tests compare whether each model improves fit compared with
the previous model.

```{r}
model_comparison <- anova(null_model, model_1, model_2, model_3)

kable(model_comparison, digits = 3)
```

# Final Model Results

Model 3 is the main hypothesis-testing model. The wave terms test H1,
loneliness tests H2, social support tests H3, and the interaction tests
H4.

```{r}
summary(model_3)
```

# Wave Comparisons

Estimated marginal means are used to compare depressive symptoms across
waves, including the key comparison between Wave 1 and Wave 3.

```{r}
wave_comparisons <- emmeans(model_3, pairwise ~ wave)

wave_comparisons
```

# Moderation Check

This checks whether the association between loneliness and depression
differs at lower, average, and higher levels of social support.

```{r}
support_sd <- sd(model_data$social_support_c, na.rm = TRUE)

moderation_check <- emtrends(
  model_3,
  ~ social_support_c,
  var = "loneliness_c",
  at = list(social_support_c = c(-support_sd, 0, support_sd))
)

moderation_check
```

# Estimation Checks

This section checks whether the final model produced a singular fit. A
singular fit can mean that the random-effects structure is too complex
for the data.

```{r}
estimation_check <- data.frame(
  model = "model_3",
  singular_fit = isSingular(model_3)
)

kable(estimation_check)
```

# Model Figures

The first figure shows model-estimated depressive symptoms across the
three survey waves. This visualises the wave effect tested in the final
mixed-effects model.

```{r}
wave_emmeans <- as.data.frame(emmeans(model_3, ~ wave))

wave_plot <- ggplot(wave_emmeans, aes(x = wave, y = emmean, group = 1)) +
  geom_line(colour = "#2F6B84", linewidth = 1) +
  geom_point(colour = "#2F6B84", size = 3) +
  geom_errorbar(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    width = 0.1,
    colour = "#2F6B84"
  ) +
  labs(
    title = "Estimated depressive symptoms across waves",
    x = "Survey wave",
    y = "Estimated depression score"
  ) +
  theme_minimal(base_size = 12)

wave_plot

ggsave(
  filename = file.path("..", "figures", "model_estimated_depression_by_wave.png"),
  plot = wave_plot,
  width = 7,
  height = 4,
  dpi = 300
)
```

The second figure shows the predicted association between loneliness and
depressive symptoms at lower, average, and higher levels of social
support. This visualises the interaction tested in the final
mixed-effects model.

```{r}
support_sd <- sd(model_data$social_support_c, na.rm = TRUE)

model_frame <- model.frame(model_3)

prediction_data <- expand.grid(
  loneliness_c = seq(
    min(model_data$loneliness_c, na.rm = TRUE),
    max(model_data$loneliness_c, na.rm = TRUE),
    length.out = 100
  ),
  social_support_c = c(-support_sd, 0, support_sd)
)

prediction_data <- prediction_data |>
  mutate(
    wave = factor("Wave 1", levels = levels(model_frame$wave)),
    sex = factor(levels(model_frame$sex)[1], levels = levels(model_frame$sex)),
    cohort = factor(levels(model_frame$cohort)[1], levels = levels(model_frame$cohort)),
    support_level = factor(
      social_support_c,
      levels = c(-support_sd, 0, support_sd),
      labels = c("Low social support", "Average social support", "High social support")
    )
  )

prediction_data$predicted_depression <- predict(
  model_3,
  newdata = prediction_data,
  re.form = NA,
  allow.new.levels = TRUE
)

moderation_plot <- ggplot(
  prediction_data,
  aes(x = loneliness_c, y = predicted_depression, colour = support_level)
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Loneliness and depressive symptoms by social support",
    x = "Loneliness score, mean-centred",
    y = "Predicted depression score",
    colour = "Social support"
  ) +
  theme_minimal(base_size = 12)

moderation_plot

ggsave(
  filename = file.path("..", "figures", "loneliness_social_support_interaction.png"),
  plot = moderation_plot,
  width = 7,
  height = 4,
  dpi = 300
)
```

# Save Model Outputs

This section saves the main model tables to the `outputs/` folder so
they can be reused in the final report.

```{r}
output_folder <- file.path("..", "outputs")

dir.create(output_folder, showWarnings = FALSE)

write_csv(model_sample_table, file.path(output_folder, "model_sample_table.csv"))
write_csv(icc_table, file.path(output_folder, "icc_table.csv"))
write_csv(as.data.frame(model_comparison), file.path(output_folder, "model_comparison.csv"))
write_csv(as.data.frame(wave_comparisons$contrasts), file.path(output_folder, "wave_comparisons.csv"))
write_csv(as.data.frame(moderation_check), file.path(output_folder, "moderation_check.csv"))
```
